---
title: "Assignment 2: Coding Basics"
author: "Sierra Kindley"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A02_CodingBasics.Rmd") prior to submission.

The completed exercise is due on Tuesday, January 21 at 1:00 pm.


## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}

sequence1 <- seq(1, 100, by = 4) #create a sequence from 1 to 100 (beginning with 1) increasing by 
#fours and assign the name 'sequence1' to said sequence

mean(sequence1); median(sequence1) 
#compute the mean of the sequence 'sequence1'; compute the median of the sequence 'sequence1'

mean(sequence1) > median(sequence1) 
#ask R if the mean of 'sequence1' is greater than the median of 'sequence1'

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}

students <- c("Anna", "John", "Katie", "Zach") #create a four-component character vector of names of 
#students and assign the name 'students' to said vector
scores <- c(84, 79, 90, 48) #create a four-component numeric vector of student test scores and assign 
#the name 'scores'to said vector
passed <- c(TRUE, TRUE, TRUE, FALSE) #create a four-component logical vector denoting whether or not 
#each student's test score was a passing grade of 50 or above
testresults <- data.frame(students, scores, passed) #combine the vectors 'students', 'scores', and 
#'passed' into a data frame and assign the name 'testresults' to said data frame
names(testresults) <- c("Student Name", "Test Score", "Passed") #label the columns of the data frame 
#'testresults' with the titles 'Student Name', 'Test Score', and 'Passed'
testresults #generate an output of the data frame 'testresults' in the console

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: This data frame is different from a matrix in that it is comprised of columns of different data types (character, numeric, logical). In contrast, a matrix contains elements that are the same type of data.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}

passtest <- function(x) {
  result <- ifelse(x >= 50, "Pass", "Fail") 
  print(result)
} #create a function with the name 'passtest' that determines whether a test score is a passing grade 
#of 50 or above; all test scores greater than or equal to 50 print "Pass", and all scores less than 
#50 print "Fail"
passtest(scores) #apply the function 'passtest' to the vector 'scores' and print the results of the 
#function 'passtest' as applied to each of the elements in the vector 'scores' in the console

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The option of writing a function using the 'ifelse' statement worked. I believe this is because using the 'ifelse' statement allows you to evaluate the entire vector based on the given conditions and print a result for each of the elements in the vector at once. If 'if' and 'else' statements were used alone, R would be unable to evaluate and generate an output for each of the elements in the vector all at once. When I tried to use 'if' and 'else' statements to write my function, I received a warning message in the console saying the condition had a length > 1 and that only the first element would be used. As they were written, the 'if' and 'else' statements would only be able to evaluate and print a result for one of the elements in the vector at a time (only the first element in the vector would be used).
