---
title: "Assignment 4: Data Wrangling"
author: "Sierra Kindley"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A04_DataWrangling.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 4 at 1:00 pm.

## Set up your session 
1. Check your working directory, load the `tidyverse` and `lubridate` packages, and upload all four raw data files associated with the EPA Air dataset. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Explore the dimensions, column names, and structure of the datasets. 

```{r, message = FALSE}
#1 
getwd()
library(tidyverse) #load 'tidyverse' package
library(lubridate) #load 'lubridate' package
EPA.O3.data2018 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv")
EPA.O3.data2019 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv")
EPA.PM25.data2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv")
EPA.PM25.data2019 <- read.csv("./Data/Raw/EPAair_PM25_NC2019_raw.csv")
#load each of the four raw data files associated with the EPA Air dataset

#2
dim(EPA.O3.data2018) #call up dimentions of 'EPA.O3.data2018' dataset
colnames(EPA.O3.data2018) #call up column names of 'EPA.O3.data2018' dataset
str(EPA.O3.data2018) #call up structure of 'EPA.O3.data2018' dataset

dim(EPA.O3.data2019) #call up dimentions of 'EPA.O3.data2019' dataset
colnames(EPA.O3.data2019) #call up column names of 'EPA.O3.data2019' dataset
str(EPA.O3.data2019) #call up structure of 'EPA.O3.data2019' dataset

dim(EPA.PM25.data2018) #call up dimentions of 'EPA.PM25.data2018' dataset
colnames(EPA.PM25.data2018) #call up column names of 'EPA.PM25.data2018' dataset
str(EPA.PM25.data2018) #call up structure of 'EPA.PM25.data2018' dataset

dim(EPA.PM25.data2019) #call up dimentions of 'EPA.OPM25.data2019' dataset
colnames(EPA.PM25.data2019) #call up column names of 'EPA.PM25.data2019' dataset
str(EPA.PM25.data2019) #call up structure of 'EPA.PM25.data2019' dataset

```

## Wrangle individual datasets to create processed files.
3. Change date to date
4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3
class(EPA.O3.data2018$Date) #check class of 'Date' column in 'EPA.O3.data2018' dataset
EPA.O3.data2018$Date <- as.Date(EPA.O3.data2018$Date, format = "%m/%d/%Y") 
#format 'Date' column as date

class(EPA.O3.data2019$Date) #check class of 'Date' column in 'EPA.O3.data2019' dataset
EPA.O3.data2019$Date <- as.Date(EPA.O3.data2019$Date, format = "%m/%d/%Y") 
#format 'Date' column as date

class(EPA.PM25.data2018$Date) #check class of 'Date' column in 'EPA.PM25.data2018' dataset
EPA.PM25.data2018$Date <- as.Date(EPA.PM25.data2018$Date, format = "%m/%d/%Y") 
#format 'Date' column as date

class(EPA.PM25.data2019$Date) #check class of 'Date' column in 'EPA.PM25.data2019' dataset
EPA.PM25.data2019$Date <- as.Date(EPA.PM25.data2019$Date, format = "%m/%d/%Y") 
#format 'Date' column as date

#4
EPA.O3.data2018 <- select(EPA.O3.data2018, Date, DAILY_AQI_VALUE, Site.Name, 
                          AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
EPA.O3.data2019 <- select(EPA.O3.data2019, Date, DAILY_AQI_VALUE, Site.Name, 
                          AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
EPA.PM25.data2018 <- select(EPA.PM25.data2018, Date, DAILY_AQI_VALUE, Site.Name, 
                            AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
EPA.PM25.data2019 <- select(EPA.PM25.data2019, Date, DAILY_AQI_VALUE, Site.Name, 
                            AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
#select the same set of columns in each of the four datasets

#5
EPA.PM25.data2018$AQS_PARAMETER_DESC = "PM2.5"
EPA.PM25.data2019$AQS_PARAMETER_DESC = "PM2.5"
#fill all cells in 'AQS_PARAMETER_DESC' column with "PM2.5" in the two PM2.5 datasets

#6
write.csv(EPA.O3.data2018, row.names = FALSE, file = "./Data/Processed/EPAair_O3_NC2018_processed.csv")
write.csv(EPA.O3.data2019, row.names = FALSE, file = "./Data/Processed/EPAair_O3_NC2019_processed.csv")
write.csv(EPA.PM25.data2018, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_NC2018_processed.csv")
write.csv(EPA.PM25.data2019, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_NC2019_processed.csv")
#save all four processed datasets in the 'Processed' folder with new file names

```


## Combine datasets
7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code. 
8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

* Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels)
* Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site, aqs parameter, and county. Take the mean of the AQI value, latitude, and longitude.
* Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
* Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
10. Call up the dimensions of your new tidy dataset.
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1718_Processed.csv"

```{r}
#7
EPA.air.data1819 <- rbind(EPA.O3.data2018, EPA.O3.data2019, EPA.PM25.data2018, EPA.PM25.data2019)
#combine the four datasets into a single new data frame

#8
EPA.air.data1819.common <- #create new dataset named 'EPA.air.data1819.common'
  EPA.air.data1819 %>%
  filter(Site.Name == "Linville Falls" | Site.Name == "Durham Armory" | 
           Site.Name == "Leggett" | Site.Name == "Hattie Avenue" | 
           Site.Name == "Clemmons Middle" | Site.Name == "Mendenhall School" | 
           Site.Name == "Frying Pan Mountain" | Site.Name == "West Johnston Co." | 
           Site.Name == "Garinger High School" | Site.Name == "Castle Hayne" | 
           Site.Name == "Pitt Agri. Center" |  Site.Name == "Bryson City" | 
           Site.Name == "Millbrook School") %>%
  #filter data to include all sites that the four datasets have in common
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  #group data by date, site, AQS parameter, and county
  summarise(meanAQI = mean(DAILY_AQI_VALUE), 
            meanLat = mean(SITE_LATITUDE), 
            meanLong = mean(SITE_LONGITUDE)) %>%
  #calculate the daily mean of the AQI value, latitude, and longitude
  mutate(month = month(Date), year = year(Date)) 
  #add columns for 'month' and 'year' by parsing 'Date' column
dim(EPA.air.data1819.common) #check dimensions of new dataset

#9
EPA.air.data1819.spread <- spread(EPA.air.data1819.common, AQS_PARAMETER_DESC, meanAQI)
#spread dataset such that AQI values for Ozone and PM2.5 are in separate columns

#10
dim(EPA.air.data1819.spread)
#call up dimensions of new dataset

#11
write.csv(EPA.air.data1819.spread, row.names = FALSE, 
          file = "./Data/Processed/EPAair_O3_PM25_NC1718_Processed.csv")
#save new processed dataset in the 'Processed' folder

```


## Generate summary tables
12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where a month and year are not available (use the function `drop_na` in your pipe). 

13. Call up the dimensions of the summary dataset.
```{r}
#12a and 12b
EPA.air.data1819.summary <- 
  EPA.air.data1819.spread %>%
  group_by(Site.Name, month, year) %>% #group data by site, month, and year
  summarise(meanAQI_O3 = mean(Ozone), 
            meanAQI_PM25 = mean(PM2.5)) %>%
  #generate mean AQI values for Ozone and PM2.5 for each group
  drop_na(month, year) #remove instances where a month and year are not available

#13
dim(EPA.air.data1819.summary)
#call up dimensions of summary dataset

```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: The 'drop_na' function allows you to specify columns in which you would like to look for NAs and drop/remove all rows that contain an NA in those specified columns. The 'na.omit' function drops/removes all rows that contain an NA in any column--it does not allow you to specify columns in which you would only like to look for an NA.