---
title: "Assignment 6: GLMs week 1 (t-test and ANOVA)"
author: "Sierra Kindley"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on t-tests and ANOVAs. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A06_GLMs_Week1.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 18 at 1:00 pm.

## Set up your session

1. Check your working directory, load the `tidyverse`, `cowplot`, and `agricolae` packages, and import the NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv dataset. 

2. Change the date column to a date format. Call up `head` of this column to verify.


```{r, message = FALSE}
#1
getwd() #verify working directory
library(tidyverse) #load tidyverse packages
library(cowplot) #load cowplot package
library(agricolae) #load agricolae package
PeterPaul.nutrients <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv")
  #import the NTL-LTER lake nutrients processed dataset

#2
class(PeterPaul.nutrients$sampledate) #check class of 'sampledate' column
PeterPaul.nutrients$sampledate <- 
  as.Date(PeterPaul.nutrients$sampledate, format = "%Y-%m-%d") 
  #format 'sampledate' column as date
head(PeterPaul.nutrients$sampledate) 
#call up head of 'sampledate' column to verify date formatting 

```


## Wrangle your data

3. Wrangle your dataset so that it contains only surface depths and only the years 1993-1996, inclusive. Set month as a factor.
```{r}

PeterPaul.nutrients <- filter(PeterPaul.nutrients, 
                                depth < 0.25 &
                                year4 > 1992 & year4 < 1997)
  #select only surface depth data from years 1993-1996
PeterPaul.nutrients$month <- as.factor(PeterPaul.nutrients$month) #set month as a factor

```

## Analysis

Peter Lake was manipulated with additions of nitrogen and phosphorus over the years 1993-1996 in an effort to assess the impacts of eutrophication in lakes. You are tasked with finding out if nutrients are significantly higher in Peter Lake than Paul Lake, and if these potential differences in nutrients vary seasonally (use month as a factor to represent seasonality). Run two separate tests for TN and TP.

4. Which application of the GLM will you use (t-test, one-way ANOVA, two-way ANOVA with main effects, or two-way ANOVA with interaction effects)? Justify your choice.

> Answer: I will use a two-way ANOVA with interaction effects, as we are examining the effects of two categorical explanatory variables (lake name and month) on a continuous response variable (TN and TP). We want to know how TN and TP (separately) vary across lakes (location) and months (seasonality). I expect the effects of lake name and month may be dependent on one another, so I have chosen to run a two-way ANOVA with interaction effects.

5. Run your test for TN. Include examination of groupings and consider interaction effects, if relevant.

6. Run your test for TP. Include examination of groupings and consider interaction effects, if relevant.

```{r}
#5
sierratheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
#build new theme 'sierratheme' and define its parameters
theme_set(sierratheme) #set sierratheme as default plot theme

# Format as aov
TN.anova.2way <- aov(data = PeterPaul.nutrients, tn_ug ~ lakename * month)
summary(TN.anova.2way)

# Run a post-hoc test for pairwise differences
TukeyHSD(TN.anova.2way)

TN.interaction <- with(PeterPaul.nutrients, interaction(lakename, month))
TN.anova.2way2 <- aov(data = PeterPaul.nutrients, tn_ug ~ TN.interaction)

TN.groups <- HSD.test(TN.anova.2way2, "TN.interaction", group = TRUE)
TN.groups

#6

# Format as aov
TP.anova.2way <- aov(data = PeterPaul.nutrients, tp_ug ~ lakename * month) 
summary(TP.anova.2way)

# Run a post-hoc test for pairwise differences
TukeyHSD(TP.anova.2way)

TP.interaction <- with(PeterPaul.nutrients, interaction(lakename, month))
TP.anova.2way2 <- aov(data = PeterPaul.nutrients, tp_ug ~ TP.interaction)

TP.groups <- HSD.test(TP.anova.2way2, "TP.interaction", group = TRUE)
TP.groups

```

7. Create two plots, with TN (plot 1) or TP (plot 2) as the response variable and month and lake as the predictor variables. Hint: you may use some of the code you used for your visualization assignment. Assign groupings with letters, as determined from your tests. Adjust your axes, aesthetics, and color palettes in accordance with best data visualization practices.

8. Combine your plots with cowplot, with a common legend at the top and the two graphs stacked vertically. Your x axes should be formatted with the same breaks, such that you can remove the title and text of the top legend and retain just the bottom legend. 

```{r}
#7
TN.anova.plot <- ggplot(PeterPaul.nutrients, aes(y = tn_ug, x = month, color = lakename)) +
  geom_boxplot() +
  xlab(expression(paste("Month"))) + #change x-label
  ylab(expression(paste("TN (", mu, "g/L)"))) + #change y-label
  labs(color = "Lake Name") +
  scale_color_manual(values = c("#0c2c84", "#7fcdbb")) + #assign colors to lakes
    stat_summary(geom = "text", fun.y = max, position = position_dodge(width = 0.75, 
                  preserve = "total"), vjust = -1, size = 3.5, 
                  label = c("ab","b","ab","b","a","b","a","b","ab","ab"))
    #add group labels to plot
print(TN.anova.plot)

TP.anova.plot <- ggplot(PeterPaul.nutrients, aes(y = tp_ug, x = month, color = lakename)) +
  geom_boxplot() +
  xlab(expression(paste("Month"))) + #change x-label
  ylab(expression(paste("TP (", mu, "g/L)"))) + #change y-label
  labs(color = "Lake Name") +
  scale_color_manual(values = c("#0c2c84", "#7fcdbb")) + #assign colors to lakes
    stat_summary(geom = "text", fun.y = max, position = position_dodge(width = 0.75, 
                  preserve = "total"), vjust = -1, size = 3.5, 
                  label = c("bcd","cd","ab","d","a","d","abc","d","abcd","cd"))
    #add group labels to plot
print(TP.anova.plot)

#8
plot_grid(TN.anova.plot, TP.anova.plot + theme(legend.position = "none"), nrow = 2, 
          align = 'v', rel_heights = c(1.25, 1))
#combine two previous graphs into one cowplot with one legend and vertically aligned axes

```




